---
title: "Prepare and explore data"
format:
  html:
    toc: true
    toc-location: right
    toc-depth: 5
    number-sections: false
    code-fold: false
    sidebar: true
    collapse-level: 4
editor: visual
editor_options: 
  chunk_output_type: console
knitr:
  opts_chunk: 
    message: false
    warning: false
    code-fold: false
    include: true
    collapse: true
    eval: true
    fig.show: hold
---


## Introduction

The main question is:

Can we identify subgroups of individuals based only on gene-expression profiles, and do those subgroups differ in clinical characteristics?

Follow-up questions: 

- What characterizes the discovered clusters?
- Do clusters differ in glucose, HbA1c, lipids, age, or other clinical variables?
- Are some clusters enriched for individuals with BMI ≥ 30?
- Which genes or clinical features contribute most to the separation between clusters?


## Load packages

```{r}
#| label: load-packages

# load packages
rm(list=ls())
library(tidyverse)

```

## Load data
```{r}
#| label: simulate-data
#| include: false
#| eval: false

rm(list=ls())
library(tidyverse)

# library(faraway)
# data(diabetes)
# head(diabetes)
# data_faraway <- diabetes
# write_csv(data_faraway, "data/data-faraway-diabetes.csv")

# Load the data
input_diabetes <- read_csv("data/data-faraway-diabetes.csv") # data diabetes from faraway package

genes_hsa <- read.csv("data/counts_anno.csv") %>% # gene annotations from a RNA-seq project
  filter(hgnc_symbol != "") %>%
  pull(hgnc_symbol)

# Calculate median BMI to replace (temporary missing values)
BMI_median <- median(input_diabetes$weight / (input_diabetes$height^2) * 703, na.rm = TRUE)

# Create BMI variable
data_diabetes <- input_diabetes %>%
  mutate(BMI = round(weight / height^2 * 703, 2)) %>%
  relocate(BMI, .after = id) %>%
  mutate(BMI = replace_na(BMI, BMI_median)) %>%
  mutate(`BMI>30` = cut(BMI, breaks = c(0, 29.9, 100), labels = c("0", "1"))) %>%
  mutate(`BMI>30` = factor(`BMI>30`, levels = c(0, 1))) |>
  relocate(`BMI>30`, .after = BMI)

# Add simulated gene expression data
data <- data_diabetes

# Set parameters
set.seed(123)
n_samples <- nrow(data)
n_genes <- 1000
n_diff <- 330

# Select some known obesity genes to include in the differentially expressed set
# obesity_genes <- c("FTO", "MC4R", "LEP", "LEPR", "POMC", "PCSK1", "BDNF", "TMEM18", "NEGR1", "SH2B1")
obesity_genes <- c("FTO", "MC4R", "LEP", "LEPR", "POMC", "PCSK1")

# Select additional DE genes
n_additional <- n_diff - length(obesity_genes)
other_diff_genes <- sample(setdiff(genes_hsa, obesity_genes), n_additional)
diff_gene_names <- c(obesity_genes, other_diff_genes)

# Remaining non-differentially expressed genes
non_diff_gene_names <- sample(setdiff(genes_hsa, diff_gene_names), n_genes - n_diff)

# All genes
all_gene_names <- sample(c(diff_gene_names, non_diff_gene_names), n_genes)

# Simulate expression matrix (normal distribution)
sim_matrix <- matrix(rnorm(n_samples * n_genes, mean = 0, sd = 1), 
                     nrow = n_samples, ncol = n_genes)
colnames(sim_matrix) <- all_gene_names

# Add differential expression for selected genes
# for (gene in diff_gene_names) {
#   idx <- which(colnames(sim_matrix) == gene)
#   sim_matrix[data$`BMI>30` == 1, idx] <- rnorm(sum(data$`BMI>30`== 1), mean = 0.3, sd = 1)
#   sim_matrix[data$`BMI>30` == 0, idx]  <- rnorm(sum(data$`BMI>30` == 0), mean = 0, sd = 1)
# }

# for (gene in obesity_genes) {
#   idx <- which(colnames(sim_matrix) == gene)
#   sim_matrix[data$`BMI>30` == 1, idx] <- rnorm(sum(data$`BMI>30` == 1), mean = 0.7, sd = 1)
#   sim_matrix[data$`BMI>30` == 0, idx]  <- rnorm(sum(data$`BMI>30` == 0), mean = 0, sd = 1)
# }


##

# latent expression signal related to BMI>30
latent_signal <- ifelse(data$`BMI>30` == 1, 1, 0) +
  rnorm(n_samples, 0, 0.9)

# moderate-effect genes
for (gene in other_diff_genes) {
  idx <- which(colnames(sim_matrix) == gene)

  sim_matrix[, idx] <-
    0.5 * latent_signal +
    rnorm(n_samples, 0, 1)
}

# stronger obesity-related genes
for (gene in obesity_genes) {
  idx <- which(colnames(sim_matrix) == gene)

  sim_matrix[, idx] <-
    0.9 * latent_signal +
    rnorm(n_samples, 0, 1)
}

##

# Convert to data frame and combine with original data
gene_df <- as.data.frame(sim_matrix)
gene_df <- gene_df %>%
  as_tibble() %>%
  mutate(id = data$id) %>%
  relocate(id, .before = everything())

# Reintroduce missing data
idx <- which(data$BMI == BMI_median)
data$BMI[idx] <- NA
data$`BMI>30`[idx] <- NA


# Save to CSV if needed
write_csv(data, "data/data-diabetes.csv")
write_csv(gene_df, "data/data-diabetes-genes.csv")

```

```{r}
#| label: load-data

# Load the clinical data
data_diabetes <- read_csv("data/data-diabetes.csv") |>
    mutate(`BMI>30` = factor(`BMI>30`))

# Load gene expression data
data_expr <- read_csv("data/data-diabetes-genes.csv")

# preview clinical data
dim(data_diabetes)
glimpse(data_diabetes)

# From the glimpse output
# - we can see that we have a mixture of numerical and categorical variables
# - and we note that we have missing data, NAs

# Check how many in each group
table(data_diabetes$`BMI>30`)

# preview gene expression data
print(dim(data_expr))
print(data_expr[1:5, 1:5])

```

## Explore data

```{r}
#| label: explore-data

# run basic EDA
# - let's look at the distribution of numerical variables
# - missing data
# - and correlation between clinical variables

data_diabetes %>%
  select("BMI>30", where(is.numeric), -id) %>% 
  pivot_longer(-`BMI>30`, names_to = "variable", values_to = "value") %>% 
  filter(is.finite(value)) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "blue4", color = "white") +
  facet_wrap(~ variable, scales = "free") +
  theme_bw() +
  labs(title = "Distribution of Numerical Variables") + 
  ylab("")

# stratify by BMI>30
data_diabetes %>%
  select("BMI>30", where(is.numeric), -id) %>% 
  pivot_longer(-`BMI>30`, names_to = "variable", values_to = "value") %>% 
  filter(is.finite(value)) %>%
  ggplot(aes(x = value, fill = `BMI>30`)) +
  geom_histogram(bins = 30, color = "white", position = "identity", alpha = 0.5) +
  facet_wrap(~ variable, scales = "free") +
  theme_bw() +
  labs(title = "Distribution of Numerical Variables") + 
  ylab("") + 
  scale_fill_manual(values = c("0" = "blue4", "1" = "red4")) 

# summarize categorical variables by obesity status
data_diabetes %>%
  select("BMI>30", where(is.character), -id) %>%
  pivot_longer(-`BMI>30`, names_to = "variable", values_to = "value") %>%
  group_by(variable, value, `BMI>30`) %>%
  summarise(count = n(), .groups = "drop") %>%
  ggplot(aes(x = value, y = count, fill = `BMI>30`)) +
  geom_col(position = "dodge") +
  facet_wrap(~ variable, scales = "free") +
  theme_bw() +
  labs(title = "Counts of Categorical Variables by Obesity Status") +
  ylab("") +
  scale_fill_manual(values = c("0" = "blue4", "1" = "red4"))

# distribution of randomly selected 10 genes
data_expr %>%
  select(id, sample(2:ncol(data_expr), 12)) %>%
  pivot_longer(-id, names_to = "gene", values_to = "expression") %>%
  ggplot(aes(x = expression)) +
  geom_histogram(bins = 30, fill = "blue4", color = "white") +
  facet_wrap(~ gene, scales = "free") +
  theme_bw() +
  labs(title = "Distribution of Randomly Selected Genes") + 
  xlab("")

```

```{r}
#| label: explore-missing-data
#| fig-cap: 
#| - "Number of missing data per variable, shows that bp.2d and bp.2s have more than 50% missing entries"

# calculate number of missing data per variable
data_na <- data_diabetes %>% 
  summarise(across(everything(), ~ sum(is.na(.)))) 

# make a table with counts sorted from highest to lowest
data_na_long <- data_na %>%
  pivot_longer(-id, names_to = "variable", values_to = "count") %>%
  arrange(desc(count)) 

# make a column plot to visualize the counts
data_na_long %>%
  ggplot(aes(x = variable, y = count)) + 
  geom_col(fill = "blue4") + 
  xlab("") + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

# Based on the number of missing data, let's delete bp.2s, bp.2d
# and use complete-cases analysis 
data_diabetes <- data_diabetes %>%
  dplyr::select(-bp.2s, -bp.2d) %>%
  na.omit()

```

## PCA

```{r}
#| label: pca
#| echo: true
#| code-fold: true
#| code-summary: "Example code"
#| eval: true

# Perform PCA on gene expression data

# filter gene expression data to include only those with obesity status
x <- data_expr %>%
  filter(id %in% data_diabetes$id) %>%
  column_to_rownames("id") %>%
  as.matrix()

# Scale the data prior pca
x_scaled <- scale(x)

# Calculate PCA
pca <- prcomp(x, center=TRUE, scale.=FALSE)
eigs <- pca$sdev^2
var_exp <- eigs / sum(eigs)

res_pca <- data.frame(PC1=pca$x[,1], PC2=pca$x[,2], PC3=pca$x[,3], PC4=pca$x[,4], PC5=pca$x[,5]) |>
    rownames_to_column("id") |> 
    mutate(id = as.integer(id)) |>
    as_tibble() |> 
    left_join(data_diabetes, by="id")

res_pca_loadings <- pca$rotation

# show top 10 loadings along PC1
res_pca_loadings |> 
    as.data.frame() |> 
    rownames_to_column("gene") |> 
    arrange(desc(abs(PC1))) |> 
    head(10) |> 
    ggplot(aes(x=reorder(gene, PC1), y=PC1)) +
    geom_bar(stat="identity", fill="steelblue") +
    coord_flip() +
    labs(title="Top genes contributing to PC1", x="gene", y="Loading") +
    theme_minimal()


# show PCA scores plot
res_pca |>
    ggplot(aes(x=PC1, y=PC2, group = `BMI>30`, fill = `BMI>30`)) +
    geom_point(alpha = 0.6,  shape = 21, size = 2) +
    labs(title="PCA scores plot", x="PC1", y="PC2") +
    xlab(paste("PC1 (Var: ", round(var_exp[1] * 100, 2), "%)")) +
    ylab(paste("PC2 (Var: ", round(var_exp[2] * 100, 2), "%)")) +
    theme_minimal() +
    theme(legend.title=element_blank()) + 
  scale_fill_manual(values = c("0" = "blue4", "1" = "red4")) 

```


```{r}

# choose how many PCs to retain for clustering
var_exp <- pca$sdev^2 / sum(pca$sdev^2)
cum_var <- cumsum(var_exp)

plot(
  cumsum(var_exp[1:400]),
  type = "l",
  xlab = "Number of PCs",
  ylab = "Cumulative variance explained"
)

which(cum_var >= 0.5)[1]
which(cum_var >= 0.7)[1]

# This curve suggests there is no sharp elbow, variance is spread fairly broadly across many components.
```

## Clustering

```{r}
#| label: cluster-data

# Use first 30 PCs for clustering for practical purposes only
x_clust <- pca$x[, 1:30]

# K-means clustering
set.seed(123)

km <- kmeans(
  x_clust,
  centers = 3,
  nstart = 50
)

# Add cluster labels to PCA results
res_pca$cluster <- factor(km$cluster)

# Plot clusters on PC1 vs PC2
res_pca %>%
  ggplot(aes(x = PC1, y = PC2, fill = cluster)) +
  geom_point(
    shape = 21,
    size = 2,
    alpha = 0.7
  ) +
  theme_minimal() +
  labs(
    title = "K-means clusters",
    x = paste0("PC1 (", round(var_exp[1] * 100, 2), "%)"),
    y = paste0("PC2 (", round(var_exp[2] * 100, 2), "%)")
  )


# Add cluster assignment to PCA results
res_pca$cluster <- factor(km$cluster)

# Export cluster results
dir.create("results", showWarnings = FALSE)

saveRDS(
  res_pca,
  "results/res_pca.rds"
)
```

## Explore clusters

We can look into how clinical variables are distributed across the clusters
```{r}

# numerical variables, e.g. BMI
res_pca |>
  ggplot(aes(x = cluster, y = BMI, color = cluster)) +
  geom_jitter() + 
  geom_boxplot(alpha = 0.5) + 
  theme_bw()

# categorical variables, e.g. frame
res_pca |>
  count(cluster, frame) |>
  group_by(cluster) |>
  mutate(prop = n / sum(n)) |>
  ggplot(aes(x = cluster, y = prop, fill = frame)) +
  geom_col(position = "stack") +
  theme_bw() +
  labs(
    x = "Cluster",
    y = "Proportion",
    fill = "Frame"
  )

```


```{r}
# Add cluster assignments to gene-expression data
expr_clusters <- data_expr |>
  inner_join(
    res_pca |> select(id, cluster),
    by = "id"
  )

# Calculate genes characterising each cluster
gene_cluster_stats <- expr_clusters |>
  pivot_longer(
    -c(id, cluster),
    names_to = "gene",
    values_to = "expression"
  ) |>
  group_by(gene) |>
  mutate(
    mean_all = mean(expression),
    sd_all = sd(expression)
  ) |>
  group_by(cluster, gene) |>
  summarise(
    mean_cluster = mean(expression),
    mean_all = first(mean_all),
    sd_all = first(sd_all),
    .groups = "drop"
  ) |>
  mutate(
    difference = (mean_cluster - mean_all) / sd_all
  )

saveRDS(
  list(
    res_pca = res_pca,
    gene_cluster_stats = gene_cluster_stats
  ),
  "results/clustering_results.rds"
)
```